Choosing Color Palettes for Statistical Graphics

### 2007-11-02 pkg/paper/Palettes/CSDA.Rnw
updated Section 2 for CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@22 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[shortnames]{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
\usepackage{amsfonts,rotating}
\usepackage{paralist}

\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien
   \And Paul Murrell\\The University of Auckland}
\Plainauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

Color is an integral element of graphical displays in general, and many statistical
graphics in particular.  Furthermore, 
an increasing amount of research is being published and viewed online, making
it easier to include color in published plots.
However, selecting which colors to use for a statistical graphic is
not a trivial task.

There are three main obstacles to overcome:
\begin{compactitem}
\item
The colors should not be unappealing.  
\end{compactitem}

It is not necessary for the 
colors in a statistical plot to reflect the latest in fashion trends, 
but basic 
principles such as avoiding fully saturated colors should be
adhered to.  The requirement is not that the user should have
a degree in graphic design, but
that the software should allow the user to select colors and provide
control over basic properties such as how bright the colors are.

In order to be able to control basic perceptual features of colors,
it is necessary to work in a \emph{colorspace} that describes 
colors in terms of their perceptual properties.
This article will describe an appropriate colorspace for
selecting colors.

\begin{compactitem}
\item 
The colors in a statistical graphic
should cooperate with each other.  
\end{compactitem}

The typical purpose 
of color in a statistical graphic is to distinguish between different 
areas or symbols in the plot---to distinguish between different
groups or between different levels of a variable.  This means
that there will typically be several colors, or a \emph{palette}
of colors, used within a plot and 
that those colours should be related to each other.

Part of the solution to this problem is to work in an
appropriate colorspace, as in the previous point, so that 
motion within the colorspace and distances between colors 
have an intuitive and rational meaning.  However, 
there still remains the issue of providing basic navigation 
tools so that users can conveniently generate color
palettes.  The main part of this article will describe
algorithms for generating color palettes for common 
applications in statistical graphics.

\begin{compactitem}
\item
The colors should work everywhere.  
\end{compactitem}

In the ideal situation, colors 
should be selected so that they continue to work in any 
context.  For example, different areas of a plot should still be 
distiguishable when printed in black-and-white, or when the person
viewing the graphic is colour-blind.
This article will also briefly address these issues.

@@

To show what can be gained by selecting appropriate color schemes, we present a collection
of illustrations with typical color graphics based on HSV palettes (always in the left panels)
and more suitable HCL palettes (always in the right panels). All HSV color palettes (and some
of the data sets) are taken from recent publications in statistical journals such as
\textit{Journal of the American Statistical Association},
\textit{Journal of Computational and Graphical Statistics} or
\textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
an overview of typically colored statistical graphics and their respective pitfalls (concerning
color palette choice).
%% All graphics are in principle very useful but can be even enhanced
%% by using better color schemes.

Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
Faithful geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications
include objective functions with two arguments \citep[as in][]{color:Gneiting+Sevcikova+Percival:2006}
or (physical) measurements on a 2-dimensional grid \citep[as in][]{color:Yang+Buckley+Dudoit:2002}.
Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 2 minutes).

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde>>=
if(generate_png) {
  data("geyser", package = "MASS")
  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
  geyser2 <- cbind(geyser$duration, geyser$waiting)
  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
  plot_bkde2D <- function(x, col = NULL, xlab = "duration", ylab = "waiting time", legend = FALSE, ...) {
    if(is.null(col)) col <- rev(grey.colors(33))
    if(legend) {
      filled.contour(x$x1, x$x2, x$fhat, col = col, nlevels = length(col), xlab = xlab, ylab = ylab, ...)
    } else {
      image(x$x1, x$x2, x$fhat,
        breaks = seq(from = min(x$fhat), to = max(x$fhat), length.out = length(col) + 1),
        xlab = xlab, ylab = ylab, col = col, axes = FALSE, cex.lab = 1.5, ...)
      box()
      axis(1, cex.axis = 1.5)
      axis(2, cex.axis = 1.5)
    }
    invisible(x)
  }

  png(file = "bkde.png", height = 810, width = 840)
  par(mfrow = c(2, 2))
  plot_bkde2D(dens, col = hsv(seq(from = 2/3, to = 0, length = 33), 0.8, 0.9))
  ## plot_bkde2D(dens, col = rainbow(33, start = 1/3, end = 1, v = 0.8, s = 0.9))
  plot_bkde2D(dens, col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  plot_bkde2D(dens, col = rev(heat.colors(33)))
  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  dev.off()
}
@
\includegraphics{bkde}
\setkeys{Gin}{width=.47\textwidth}
<<bkde-pal>>=
if(generate_png) {
  png(file = "bkde-pal.png", height = 160, width = 396)
  par(mfrow = c(4, 1), mar = c(0, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    hsv(seq(from = 2/3, to = 0, length = 99), 0.8, 0.9))
    ## rainbow(99, start = 1/3, end = 1, v = 0.8, s = 0.9))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat.colors(99)))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(sequential_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  box()
  axis(1, cex.axis = 1.5)
  dev.off()
}
@
\includegraphics{bkde-pal}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics (or on presentation slides and in interactive usage in statistical
software packages), many users prefer to have colored displays and most often use HSV
palettes (as in the two left two panels). These palettes code the variable of interest
by varying hue in an HSV color wheel as done by \citet[Figure~4a]{color:Yang+Buckley+Dudoit:2002}
or \citet[Figures~1--4]{color:Gneiting+Sevcikova+Percival:2006}. The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works a bit better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{color:Ihaka:2003}.

%FIXME% examples for sequential palettes:
%%      color:Gelfand+Banerjee+Sirmans:2007

In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
coding the variable of interest much better and without flashy colors. If, however, the user wants
to add some color to the plot, this could be done by using a better balanced version of the heat
colors (as shown in the bottom right panel). These colors also increase from a yellow to a red
hue while being balanced towards the same gray as in the grayscale palette (i.e., when converted
to a grayscale or printed out on a grayscale printer, the upper and lower right panel would 
essentially look identical). Both palettes have in common that they give increasing perceptual emphasis
to regions with increasing density, resulting in a heatmap that highlights the (small)
interesting high-density regions and not to the large low-density regions surrounding them.

@@

  png(file = "nigeria.png", height = 1050, width = 840)
  par(mfrow = c(3, 2), mar = c(1.22, 0.82, 1.02, 0.82))
  plot(m, z = fit$pmode, fill = rev(rainbow(99, start = 0, end = 1/3)),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, fill = dichromat(rev(rainbow(99, start = 0, end = 1/3)), type = "protan"),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  dev.off()
}
@
\includegraphics{nigeria}
\setkeys{Gin}{width=.47\textwidth}
<<nigeria-pal>>=
if(generate_png) {
  png(file = "nigeria-pal.png", height = 120, width = 396)
  par(mfrow = c(3, 1), mar = c(0, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(rainbow(99, start = 0, end = 1/3)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  axis(1, cex = 1.5)
  dev.off()
}
@
\includegraphics{nigeria-pal}
\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.}
\end{center}
\end{figure}

A similar example is presented in Figure~\ref{fig:nigeria}
\citep[taken from][Figure~5, left]{color:Kneib:2006}, depicting posterior mode estimates
for childhood mortality in different regions of Nigeria. Deviations from a model for
childhoold mortality are brought out by shading a map according to the corresponding
residuals, revealing decreased mortality in the south-west and increased mortality in
the north-east. \cite{color:Kneib:2006} uses an HSV-based palette coding the residuals by
the hue, going from green via yellow to red. Our HCL-based palette also employs green
and red hues for negative and postive residuals respectively, but codes neutral values
(around 0) by a neutral light gray. Compared to the HSV-based palette, this offers again
a number of advantages: only the interesting areas are highlighted by high-chroma colors;
flashy colors are avoided making it easier to look at the display for a longer time; positive
and negative residuals with the same absolute size receive the same perceptual weight
by being balanced towards the same gray.

%FIXME% other bad examples for diverging
%%      color:Harezlak+Coull+Laird:2007
%%      color:Tenenhaus+Giron+Viennet:2007

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<seats, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
names(parties_hcl) <- names(parties_hsv) <- names(seats)
names(seats)[4] <- "Grüne"
pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

Although problematic for many tasks, pie charts can be useful for visualizing
whether a set of pie segments constitutes a majority. A typical application is shown
in Figure~\ref{fig:seats}, visualizing the distribution of seats in the German parliament
``Bundestag'' following the 2005 election. In this election, five parties were able to
obtain enough votes to enter the Bundestag---however, neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists ``Die Linke'', this lead to
a big coalition of CDU/CSU and SPD. In graphical displays, the parties are usually matched
by using colors as metaphors: red for the social democrats SPD, black for the conservative
CDU/CSU, yellow for the liberal FDP, green for the green party ``Die Gr\"unen'' and purple 
for the leftist party ``Die Linke''. The left panel shows fully saturated HSV colors as usually
found in the (German) media whereas the right panel uses less flashy HCL colors with the
same hues (except for the CDU/CSU where a blue hue instead of the extreme ``color'' black
is used). The advantage of the latter is again that they are easier to hold in focus for
a longer time. Furthermore, they are all balanced towards the same gray and have the same
amount of color, resulting in a perceptually balanced palette that does not introduce undesired
graphical distractions. While it could be argued that this pie chart is such a simple
display that a well-balanced palette is not so important and that flashy colors are to
be preferred, the balancing properties of the HCL-based palette are very important in more
complex displays such as the mosaic display in Figure~\ref{fig:votes}. This re-uses the same
colors and shows the distribution of votes in the 2005 election
stratified by province (Bundesland). The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

\setkeys{Gin}{width=\textwidth}
\begin{sidewaysfigure}[p]
\begin{center}
<<votes, fig=TRUE, height=6, width=14, eval=FALSE>>=
data("Bundestag2005", package = "vcd")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties_hcl <- parties_hcl[colnames(votes)]
parties_hsv <- parties_hsv[colnames(votes)]
dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hsv),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hcl),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport(2)
@
\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
\end{center}
\end{sidewaysfigure}

Color choice is usually much more important in graphical displays with
shaded areas compared to displays with only points or lines. However, color choice
becomes much more relevant also in scatter plots when there are many points. As an
example, Figure~\ref{fig:scatter} depicts a scatter plot with three clusters similar to 
the one shown in \citet[Figure~3]{color:Celeux+Hurn+Robert:2000}. To indicate cluster
membership, three different colors (green, yellow, red) are employed. The HSV colors in
the left panel are again very flashy and differ strongly with respect to luminance: the
yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
the amount of color and gray.

As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{color:Friendly:2002}
visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
Clearly, hair and eye color are not independent and the pattern of association is highlighted
by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
it is shown again that the HCL colors are better balanced (between the red and blue colors)
and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
to read when shaded in a light grey rather than white. More details and extensions to residual-based
shadings can be found in \cite{color:Zeileis+Meyer+Hornik:2007}.

>>>>

\documentclass[shortnames]{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}

%FIXME% Different/better title?
\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien
   \And Paul Murrell\\The University of Auckland}
\Plainauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

Color is an integral element of graphical displays in general, and many statistical
graphics in particular.  Furthermore, 
an increasing amount of research is being published and viewed online, making
it easier to include color in published plots.
However, selecting which colors to use for a statistical graphic is
not a trivial task. There are three main obstacles to overcome:

\paragraph{The colors should not be unappealing.}
It is not necessary for the 
colors in a statistical plot to reflect the latest in fashion trends, 
but basic 
principles such as avoiding fully saturated colors should be
adhered to.  The requirement is not that the user should have
a degree in graphic design, but
that the software should allow the user to select colors and provide
control over basic properties such as how bright the colors are.
In order to be able to control basic perceptual features of colors,
it is necessary to work in a \emph{colorspace} that describes 
colors in terms of their perceptual properties.
This article will describe an appropriate colorspace for
selecting colors.

\paragraph{The colors in a statistical graphic should cooperate with each other.}
The typical purpose 
of color in a statistical graphic is to distinguish between different 
areas or symbols in the plot---to distinguish between different
groups or between different levels of a variable.  This means
that there will typically be several colors, or a \emph{palette}
of colors, used within a plot and 
that those colours should be related to each other.
Part of the solution to this problem is to work in an
appropriate colorspace, as in the previous point, so that 
motion within the colorspace and distances between colors 
have an intuitive and rational meaning.  However, 
there still remains the issue of providing basic navigation 
tools so that users can conveniently generate color
palettes.  The main part of this article will describe
algorithms for generating color palettes for common 
applications in statistical graphics.

\paragraph{The colors should work everywhere.}
In the ideal situation, colors 
should be selected so that they continue to work in any 
context.  For example, different areas of a plot should still be 
distiguishable when printed in black-and-white, or when the person
viewing the graphic is colour-blind.
This article will also briefly address these issues.
%FIXME% Also mention that the colors should work on different devices.
%FIXME% Explain that we discuss how and under which circumstances this
%FIXME% is achieved.

%FIXME% A sentence or so is needed as a "bridge".

@@

To show what can be gained by selecting appropriate color schemes,
we present two examples for typical color graphics, contrasting
commonly-used HSV palettes with more suitable HCL palettes.
Both examples and all HSV palettes are taken from (the electronic
version of) recent publications in
\textit{Computational Statistics \& Data Analysis} (Volume~51).

Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old Faithful
geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications include 
classification maps \citep[as in][Figure~2]{color:Tenenhaus+Giron+Viennet:2007},
function surfaces \citep[as in][Figure~8]{color:Harezlak+Coull+Laird:2007}, or
measurements on a spatial grid \citep[as in][Figure~3]{color:Gelfand+Banerjee+Sirmans:2007}.
Figure~\ref{fig:bkde} shows a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 2 minutes).

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde>>=
if(generate_png) {
  data("geyser", package = "MASS")
  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
  geyser2 <- cbind(geyser$duration, geyser$waiting)
  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
  plot_bkde2D <- function(x, col = NULL, xlab = "duration", ylab = "waiting time", legend = FALSE, ...) {
    if(is.null(col)) col <- rev(gray.colors(33))
    if(legend) {
      filled.contour(x$x1, x$x2, x$fhat, col = col, nlevels = length(col), xlab = xlab, ylab = ylab, ...)
    } else {
      image(x$x1, x$x2, x$fhat,
        breaks = seq(from = min(x$fhat), to = max(x$fhat), length.out = length(col) + 1),
        xlab = xlab, ylab = ylab, col = col, axes = FALSE, cex.lab = 1.5, ...)
      box()
      axis(1, cex.axis = 1.5)
      axis(2, cex.axis = 1.5)
    }
    invisible(x)
  }

  png(file = "bkde.png", height = 810, width = 840)
  par(mfrow = c(2, 2))
  plot_bkde2D(dens, col = hsv(seq(from = 2/3, to = 0, length = 33), 0.8, 0.9))
  ## plot_bkde2D(dens, col = rainbow(33, start = 1/3, end = 1, v = 0.8, s = 0.9))
  plot_bkde2D(dens, col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  plot_bkde2D(dens, col = rev(heat.colors(33)))
  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  dev.off()
}
@
\includegraphics{bkde}
\setkeys{Gin}{width=.47\textwidth}
<<bkde-pal>>=
if(generate_png) {
  png(file = "bkde-pal.png", height = 160, width = 396)
  par(mfrow = c(4, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    hsv(seq(from = 2/3, to = 0, length = 99), 0.8, 0.9))
    ## rainbow(99, start = 1/3, end = 1, v = 0.8, s = 0.9))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat.colors(99)))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  box()
  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
    rev(sequential_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  box()
  axis(1, cex.axis = 1.8)
  dev.off()
}
@
\includegraphics{bkde-pal}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics (or on presentation slides and in interactive usage in statistical
software packages), many users prefer to have colored displays and most often use HSV
palettes (as in the two left two panels). These palettes code the variable of interest
by varying hue in an HSV color wheel as done by 
\citet[Figure~8]{color:Harezlak+Coull+Laird:2007}
or \citet[Figure~2]{color:Tenenhaus+Giron+Viennet:2007}. 
The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works a bit better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{color:Ihaka:2003}.

In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
coding the variable of interest much better and without flashy colors. If, however, the user wants
to increase the contrast by adding some color to the plot, this could be done by using a better
balanced version of the heat colors (as shown in the bottom right panel). These colors also
increase from a yellow to a red hue while being balanced towards the same gray as in the
grayscale palette. Thus, when converted to a grayscale or printed out on a grayscale printer
(as in the printed version of \textit{Computational Statistics \& Data Analysis}),
the upper and lower right panel would look (virtually) identical.
Both palettes have in common that they give increasing perceptual emphasis
to regions with increasing density, resulting in a heatmap that highlights the (small)
interesting high-density regions and not to the large low-density regions surrounding them.
By balancing the lower right panel towards light/dark contrasts, it is assured that the graphic
is intelligible for color-blind readers and that the same graphic works in
colored electronic versions and printed grayscale versions of a publication.

@@

  png(file = "nigeria.png", height = 1050, width = 840)
  par(mfrow = c(3, 2), mar = c(1.22, 0.82, 1.02, 0.82))
  plot(m, z = fit$pmode, fill = rev(rainbow(99, start = 0, end = 1/3)),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, fill = dichromat(rev(rainbow(99, start = 0, end = 1/3)), type = "protan"),
    zlim = c(-0.65, 0.65), legend = FALSE)
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
    fill = dichromat(rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)), type = "protan"))
  dev.off()
}
@
\includegraphics{nigeria}
\setkeys{Gin}{width=.47\textwidth}
<<nigeria-pal>>=
if(generate_png) {
  png(file = "nigeria-pal.png", height = 120, width = 396)
  par(mfrow = c(3, 1), mar = c(0.5, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(rainbow(99, start = 0, end = 1/3)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
    rev(diverge_hcl(99, h = c(300, 120), c = 60, l = c(45, 90), power = 1.2)))
  box()
  axis(1, cex.axis = 1.8)
  dev.off()
}
@
\includegraphics{nigeria-pal}
\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.}
\end{center}
\end{figure}

The second example is presented in Figure~\ref{fig:nigeria}
\citep[taken from][Figure~5, left]{color:Kneib:2006}, depicting posterior mode estimates
for childhood mortality in different regions of Nigeria. Spatial variations (not included
in the model) in childhood mortality are brought out by shading a map according to the
corresponding model deviations, 
revealing decreased mortality in the south-west and increased mortality in
the north-east. \cite{color:Kneib:2006} uses an HSV-based palette coding the residuals by
the hue, going from green via yellow to red (see the upper left panel of Figure~\ref{fig:nigeria}).
Similar approaches are used by \citet[Figure~8]{color:Harezlak+Coull+Laird:2007} and
\citet[Figure~2]{color:Tenenhaus+Giron+Viennet:2007}. 
Our first HCL-based palette in the middle left panel also employs green
and red hues for negative and postive residuals respectively, but codes neutral values
(around 0) by a neutral light gray. Compared to the HSV-based palette, this offers again
a number of advantages: only the interesting areas are highlighted by high-chroma colors;
flashy colors are avoided making it easier to look at the display for a longer time; positive
and negative residuals with the same absolute size receive the same perceptual weight
by being balanced towards the same gray.

Another advantage of the HCL-based palette becomes apparent when we collapse the
red-green distinctions in the colors to emulate what a person with protanopia
(a common red-green color blindness) would see. The right panel of Figure~\ref{fig:nigeria}
displays the respective palettes from the left panel, but transformed for protanopic
vision using the tools of \cite{color:Lumley:2006}. In the upper right panel, the
low-mortality regions in the south-west just ``disappear'' because they are transformed
to virtually the same yellow as the average-mortality regions. In contrast, in the middle
right panel, regions with increased and decreased mortality remain clearly distinguishable.
Only the type and amount of color is changed (red turns gray, green turns yellow/brown) but
regions with large deviations are still easily indentifiable by being balanced towards
darker grays. To avoid that one of the two ``branches'' of the palette is only gray
in the protanopic version, we could also employ a purple (rather than a red) hue for
positive deviations. This is used in the lower panels of Figure~\ref{fig:nigeria}
showing that after collapsing red-green distinctions, a useful yellow/brown--gray--blue
palette remains from the original green--gray--purple palette. Thus, the lower left
panel could be used ensuring that it works for both normal and protanopic vision.
Similar arguments hold for deuteranopic vision.
